Reliable feature point matching

### 2013-08-13 text/head/titlepage.tex
A few touches on the thesis

\null\vspace{2cm}
{\huge Reliable feature point matching} \\[24pt] 
\textcolor{gray}{\small{THIS IS A TEMPORARY TITLE PAGE \\ It will be 
replaced for the final print by a version \\ provided by the service 
academique.}}

@@

\begin{tabular} {cc}
\parbox{0.3\textwidth}{\includegraphics[width=4cm]{images/epfl}}
&
\parbox{0.7\textwidth}{%
	Thesis in Computer Science\\
	presented on September 3rd 2013\\
	at the Advanced Science Research Center\\
	Fusionopolis, Singapore\\

%
Singapore, ADSC, 2011}
\end{tabular}
\end{center}
\vspace{2cm}
\end{titlepage}

>>>>

\null\vspace{2cm}
{\huge Reliable Feature Point Matching} \\[24pt]
{\small MASTER'S THESIS}

@@

\begin{tabular} {cc}
\parbox{0.3\textwidth}{\includegraphics[width=4cm]{images/adsc}\\[18pt]
\includegraphics[width=4cm]{images/epfl}}
&
\parbox{0.7\textwidth}{%
    Jonas Toft Arnfred\\
    Master's student in Computer Science\\
	at École Polytechnique Fédérale de Lausanne (EPFL)\\[9pt]
	Spring semester 2013\\
	Advanced Digital Sciences Center (Singapore)\\[9pt]
    Supervisors at ADSC: \ Dr. Stefan Winkler\\ [4pt]
    Supervisor at EPFL: ~ \ Prof. Sabine Süsstrunk\\ [4pt]
    \null \hspace{3em} }

%
%Singapore, ADSC, August 2013
\end{tabular}
\end{center}
\vspace{2cm}
\end{titlepage}

### 2013-08-13 text/main/ch1_introduction.tex
A few touches on the thesis

My goal when I started working on this thesis was not to improve feature 
point matching, but via a series of detours I ended up circling the 
problem which attracted me, not just because it its importance, but also 
because of its inherent simplicity. Out of two sets, finding pairs of 
similar elements seems like a problem that should have been solved long 
ago, yet because of the nature of feature points and the data they are 
used on, there seems to be plenty of progress to be made.

As the outcome of my work  I propose two algorithms, \emph{Mirror Match 
(MM)} and \emph{Mirror Match with Clustering (MMC)}, both designed to 
improve feature point matching. The goal of this thesis will be to 
introduce these algorithms and convincingly demonstrate that they are 
useful. While doing this I hope to provide sufficient background for a 
technically minded reader without any necessary background in computer 
vision to follow the motivation behind \emph{MM} and \emph{MMC} as well 
as the place of both algorithms in a bigger scientific perspective.

@@

For the remainder of the thesis I will make use of the academic 
'\emph{we}' to underline the fact this thesis had not been possible had 
it not been for the guidance of Stefan Winkler and Sabine S\"usstrunk as 
well as the support of Advanced Digital Science Center in Singapore that 
hosted me as an intern during my thesis work.

>>>>

My goal when I started working on this thesis was not to improve feature 
point matching, but via a series of detours I ended up circling the 
problem which I found alluring, not just because it its importance, but 
also because of its inherent simplicity. Out of two sets, finding pairs 
of similar elements seems like a problem that should have been solved 
long ago, yet because of the nature of feature points and the data they 
are used on, there seems to be plenty of progress to be made still.

As the outcome of my work I propose two algorithms, \emph{Mirror Match 
(MM)} and \emph{Mirror Match with Clustering (MMC)}. Both algorithms are 
designed to tackle the problem of reliably matching feature points under 
circumstances where we cannot assume two images necessarily have any 
correspondences.  The goal of this thesis will be to describe these 
algorithms and convincingly demonstrate that they outperform traditional 
approaches.  While doing this I hope to provide sufficient background 
for a technically minded reader without any necessary experience in 
computer vision to follow the motivation behind \emph{MM} and \emph{MMC} 
and enable him or her to place both algorithms in a bigger scientific 
perspective.

@@

For the remainder of the thesis I will make use of the academic 
'\emph{we}' to underline the fact this thesis had not been possible had 
it not been for the guidance of \emph{Stefan Winkler} and \emph{Sabine 
S\"usstrunk} as well as the support of \emph{Advanced Digital Science 
Center} in Singapore that hosted me as an intern during the work on this 
thesis.

### 2013-08-13 text/main/ch2_background.tex
Added some introduction to matching methods
Corrections
Fixed references
A few touches on the thesis

\chapter{Matching Local Image Features}
\label{C:background}

@@

The field of computer vision is anecdotally said to have been founded 
when Marvin Minsky assigned the problem of making a computer see as an 
undergraduate summer project back in 1966\footnote{This is the same 
Marvin who invented neural networks and went on to serve as scientific 
adviser on Kubrick's movie \emph{Space Odyssey 2001}}. Whether or not 
the anecdote is true, the idea that the field of computer vision should 
be solvable in a summer's worth of time but still to this days remains 
an active field of research, illustrates an interesting notion: Making 
computers solve tasks involving seeing turns out to be much harder than 
we would initially assume.

Part of this notion stems from the fact that vision is a problem that we 
as humans are very good at. We can easily recognize objects under 
different light, or stitch together photographies to a panorama, or 
navigate a corridor identifying obstacles along the way. In fact these 
things are so easy to us, that it seems strange that they should be 
difficult at all, yet consistently recognizing an object under different 
lighting conditions and from different angles can still cause problems 
for even the best object recognition algorithms.

@@

The main factor that makes computer vision difficult is how the world we 
see around us constantly varies. This is not something that is 
immediately obvious to the casual observer, so for the sake of clarity 
let us do a small thought experiment. Try for a moment to imagine that 
you enter a room and recognize the newspaper that is lying on the table, 
illuminated from the grey light coming in of the window on the opposing 
wall. You move across the room and switch on the light make out if the 
paper is from today, and then you sit down and put on your glasses to 
read the headline. Try to imagine the newspaper as you see it on three 
different stages. First as you enter the room, then as you turn after 
turning on the light, now standing on the other side of the table, and 
last when you are sitting with the paper in your hand after you put on 
your glasses. For each of these stages, picture the situation as a 
photography and look at how the newspaper changes as you move around. At 
first you see it at an angle lying in a fairly dark and grey room, the 
white pages reflecting the light from the window with a cold bluish hue 
and the words too far away to make out. Then in the second stage with 
light on, the object you are looking at has suddenly changed. When you 
turned on the light in the room, you also changed the color of the 
newspaper, which now has a warm and yellowy white color.  The angle you 
are looking at the paper from has changed too, and the words and 
pictures are no longer upside down, although still slightly rotated and 
with the increased light in the room the contrast of the words and 
pictures of the page makes it possible for you to make recognize that 
the newspaper is from today. Finally as you sit down and put your 
glasses on the newspaper changes again. The color might be the same, but 
if you imagine taking a photography as you are sitting with the paper in 
your hand, the newspaper will look much bigger almost taking up the 
entire frame. Now that your glasses are on, the words stand out much 
clearer and you can see what is in the pictures on the front page.

Intuitively you know that the paper you sat down to read is the same 
object as the paper you saw when you entered the room, but if we limit 
ourselves to only looking at the object itself, almost everything about 
it changed in the process, from the color of the pages to the sharpness 
of the lines. On top of that the paper has changed in size and rotation 
from lying on the table to being in your hand. In our daily lives we are 
aided in the process of recognizing the paper by our memory and idea of 
how the world works. When we walked in to the dark room and saw 
something on the table, it was probably safe to assume that it was a 
newspaper. Maybe the newspaper always lies like that on the table, or 
maybe we left it there ourselves. Afterwards as we move around in the 
room, we can safely assume that the paper didn't move and even if it 
might not look exactly the same after we turned on the light, it's safe 
to guess that nobody took the old newspaper and replaced it with a new 
and yellower version.

On the other hand a computer algorithm trying to recognize objects will 
not be able to make the same assumptions about the world. We might just 
give as input to the program to pictures of the newspaper. One from the 
first stage where the light was turned off, the paper was far away and 
the lens wasn't focused correctly, and one from the third stage with the 
light on, the image in focus and the paper taking up the entire photo.  
Now we ask the algorithm: "Is this the same newspaper?". In this case it 
is fair to ask how the algorithm can reliably make this judgement.

@@

The example of the newspaper focuses on a particular branch of computer 
vision; object recognition, but the exact same problem remains for most 
of the various subfields. In automatic panorama stitching where we try 
to create a panorama from a set of images, we need to recognize which 
images fit together even if they are taken with different camera 
settings. In near duplicate detection where we try to find images that 
appear to be duplicates of other images we need to recognize if the same 
scene appears from slightly different angles under slightly different 
light. In scene matching where we try to find out if two images are 
taken of the same scene the two images could be taken on different days 
under different conditions. The list continues and it should hopefully 
be clear that in order to solve a lot of problems in computer vision, we 
need a way to deal with these variations.

@@

As Figure~\ref{fig:feature_point} illustrates, an easy heuristic for 
finding a unique point in an image is to find corners. A point lying on 
a smooth part of the image would look identical to the surrounding 
points, and points along an edge will look the same as other points 
along the same edge. While corners aren't necessarily a sufficient 
condition for finding a good feature point, they provide a good starting 
point. Intuitively the simplest way to find corners would be to detect 
edges in the image (for example using a high pass filter), and follow 
these edges until they turn. Some of these approaches are detailed in 
\cite{university1978comparison}.

A different (and it turns out, better) approach is to look at the 
gradient across the image and find spots where there is a gradient in 
two perpendicular directions.  That is, along a dark edge on a bright 
background there will be a gradient perpendicular to the edge, but no 
gradient along the edge.  However in a corner we would have a gradient 
in two different directions. The idea was originally introduced by 
Morevec \cite{hans1977towards} and later improved upon by Harris and 
Stephens who introduced what is now commonly known as the Harris corner 
detector \cite{harris1988combined}.

@@

For a thorough review of the history of feature detectors, Tuytelaars 
and Mikolajczyk presents an excellent historical retrospect in 
\cite{tuytelaars2008local}.

@@

Instead of using the pixel values, we can create a histogram of the 
gradients surrounding the feature point as Lowe did in his Scale 
Invariant Feature Detector (SIFT) \cite{lowe2004sift}. The histogram he 
proposes is based on the gradient locations and orientations weighted by
the gradient magnitude. To make sure that the histogram is rotation 
invariant, the bins of the histogram are normalized and start with the 
bin of the highest magnitude. The SIFT descriptor has been very widely 
used since it's introduction and has become a de facto standard in 
computer vision. To enhance the computational time needed to compute 
descriptors, Bay et al.\ introduced the Speeded Up Robust Features 
(SURF) that rely on a histogram of wavelets instead of gradients as used 
in SIFT \cite{bay2006surf}.

A radically different approach to a feature descriptor is to measure the
intensity difference between a number of pixels surrounding the key 
point and simply assign the bits to a vector that is returned as the 
feature descriptor. Calonder et al.\ \cite{calonder2010brief} introduced 
a feature descriptor, Binary Robust Independent Elementary Features 
(BRIEF), based on this idea. The main advantage of their approach is 
reduced processing and memory needed to create, store and match the 
feature descriptors. Interestingly the positions used for pixel wise 
comparisons were selected at random and tested against similar random 
configurations using the configuration yielding the best result. The 
simplicity of this method allows for real time matching of objects from 
a video stream, something that would be too computationally tolling when 
using a descriptor like SIFT\@. BRIEF however fails to be rotation 
invariant making it less useful in cases where images are not 
necessarily upright.

@@

Most applications of feature points use them to match features across 
different images. The standard out of the box way most feature points 
are matched involve comparing the feature descriptor using some kind of 
distance measure. If we match all features of one image with all 
features of another and discard all matches that are below a certain 
threshold of similarity, we have in essence a set of matches between the
two images. For most feature points (for example SIFT and SURF), this 
similarity is measured by the cosine similarity. Binary descriptors such
as BRIEF can take advantage of the faster computation of the hamming 
distance.

@@

An alternative is to pick out different areas in each image and pair 
areas when correspondence pairs are mainly found going from one area to 
another. This allows for the filtering of all correspondences that do 
not fall within paired areas. Examples include \emph{Isodata} 
\cite{das2008event}, which clusters feature points according to their 
position in the image, and the method from \cite{wu2011robust}, which 
uses maximally stable extremal regions (MSER) feature points to 
designate areas.

@@

In practice there are many situations where relying on the geometry of 
the image to filter correspondences is not possible, for example images 
without overlap, or adjacent objects in one image are that are separated 
in another. In these cases the assumptions behind the geometric 
algorithms will often cause them to return incorrect matches.

@@

In the absence of geometric constraints, only few methods are available.  
As pointed out in \cite{szeliski2010} and mentioned above, the simplest 
approach to matching feature points is to return all matches above a 
given threshold of similarity.  This can be done with a fixed threshold 
or by finding the nearest neighbors for each point (still filtering by a 
threshold to avoid including matches for feature points that have no 
counterpart in the other image).  When Lowe \cite{lowe2004sift} proposed 
the scale-invariant feature transform he included an alternative 
matching measure. Instead of rating matching by similarity assessed the 
uniqueness of a given match by looking at the two nearest neighbors of 
each feature point and calculating the matching score as the ratio of 
similarities. This method will later be referred to as \emph{Ratio}.  
Ranking the scores by their uniqueness and picking the $n$ best produces 
a set of correspondences that are distinctly matched across two images.  
Of these three methods (fixed threshold, nearest neighbors, and 
\emph{RATIO}), \emph{RATIO} was found to perform best by a slight margin 
\cite{mikolajczyk2005performance}.

>>>>

\chapter{Matching Local Image Features}
\label{C:Background}

@@

The field of computer vision is anecdotally said to have been founded 
when Marvin Minsky assigned the problem of making a computer see as an 
undergraduate summer project back in 1966\footnote{This is the same 
Marvin who invented neural networks and went on to serve as scientific 
adviser on Kubrick's movie \emph{Space Odyssey 2001}}. Whether or not 
the anecdote is true, the idea that the field of computer vision should 
be solvable in a summer's worth of time but still to this days remains 
an active field of research, illustrates an interesting notion: Making 
computers solve tasks involving vision turns out to be much harder than 
we would initially assume.

Part of this notion stems from the fact that vision is a problem that we 
as humans are very good at. We can easily recognize objects under 
different light, stitch together photographies to a panorama or navigate 
a corridor identifying obstacles along the way. In fact these things are 
so easy to us, that it seems strange that they should be difficult at 
all, yet consistently recognizing an object under different lighting 
conditions and from different angles can still cause problems for even 
the best object recognition algorithms.

@@

The main factor that makes computer vision difficult is how the world we 
see around us constantly varies. This is not something that is 
immediately obvious to the casual observer, so for the sake of clarity 
let us do a small thought experiment. Try for a moment to imagine that 
you enter a room and recognize the newspaper that is lying on the table, 
illuminated from the grey light coming in of the window on the opposing 
wall. You move across the room and switch on the light make out if the 
paper is from today, and then you sit down and put on your glasses to 
read the headline. Try to imagine the newspaper as you see it at three 
different stages. First as you enter the slightly obscure room, then as 
you turn around after turning on the light, now standing on the other 
side of the table, and finally when you are sitting with the paper in 
your hands after you have put on your glasses. For each of these stages, 
try to mentally picture the situation as a photography and look at how 
the newspaper changes as you move around. At first you see it at an 
angle lying in a fairly dark and grey room, the white pages reflecting 
the light from the window with a cold bluish hue and the words too far 
away to make out.  Then in the second stage with the lights turned on, 
the object you are looking at has suddenly changed.  When you turned on 
the light in the room, you also changed the color of the newspaper, 
which now has a warm and yellowy white color to its pages.  The angle 
you are looking at the paper from has changed too, and the words and 
pictures are no longer upside down, although still slightly rotated and 
with the increased light in the room the contrast of the words and 
pictures of the page makes it possible for you to make recognize that 
the newspaper is from today.  Finally as you sit down and put your 
glasses on the newspaper changes again. The color might be the same, but 
if you imagine taking a photography as you are sitting with the paper in 
your hand, the newspaper will look much bigger, almost taking up the 
entire frame. Now that your glasses are on, the words stand out much 
clearer and you can see what is in the pictures on the front page.

Intuitively you know that the paper you sat down to read is the same 
object as the paper you saw when you entered the room, but if we limit 
ourselves to only looking at the object itself, almost everything about 
it changed in the process, from the color of the pages to the sharpness 
of the lines. On top of that, the paper has changed in size and rotation 
from lying on the table to being in your hand. In our daily lives we are 
aided in the process of recognizing the paper by our memory and idea of 
how the world works. When we went in to the dark room and saw something 
on the table, it was probably safe to assume that it was a newspaper.  
Maybe the newspaper always lies like that on the table, or maybe we left 
it there ourselves. Afterwards when we move around in the room, we can 
safely assume that the paper didn't move while our back is turned and 
even if it might not look exactly the same after we turned on the light, 
it's safe to guess that nobody took the old newspaper and replaced it 
with a new and yellower version.

On the other hand a computer algorithm trying to recognize objects will 
not be able to make the same assumptions about the world. We might just 
give as input to the program two pictures of the newspaper. One from the 
first stage where the light was turned off, the paper was far away and 
the lens wasn't focused correctly, and one from the third stage with the 
light on, the image in focus and the paper taking up the entire photo.  
Now we ask the algorithm: "Is this the same newspaper?". In this case it 
is fair to ask how the algorithm can reliably make this judgement.

@@

The example of the newspaper focuses on a particular branch of computer 
vision; object recognition, but the exact same problem remains for most 
of the various subfields. In automatic panorama stitching where we try 
to create a panorama from a set of images, we need to recognize which 
images fit together even if they are taken with different camera 
settings. In near duplicate detection where we try to find images that 
appear to be duplicates of other images we need to recognize if the same 
scene appears from slightly different angles under slightly different 
light. In scene matching where we try to find out if two images are 
taken of the same scene the two images could be taken on different days 
under different conditions. The list goes on and it should hopefully be 
clear that in order to solve a lot of problems in computer vision, we 
need a way to deal with these variations.

@@

As Figure~\ref{fig:feature_point} illustrates, an easy heuristic for 
finding a unique point in an image is to find corners. A point lying on 
a smooth part of the image would look identical to the surrounding 
points, and points along an edge will look the same as other points 
along the same edge. While corners aren't necessarily a sufficient 
condition for finding a good feature point, they provide a good starting 
point.

Intuitively the simplest way to find corners would be to detect edges in 
the image (for example using a high pass filter), and follow these edges 
until they turn. Some of these approaches are detailed in 
\cite{university1978comparison}. A different (and it turns out, better) 
approach is to look at the gradient across the image and find spots 
where there is a gradient in two perpendicular directions.  That is, 
along a dark edge on a bright background there will be a gradient 
perpendicular to the edge, but no gradient along the edge.  However in a 
corner we would have a gradient in two different directions. The idea 
was originally introduced by Morevec \cite{hans1977towards} and later 
improved upon by Harris and Stephens who introduced what is now commonly 
known as the Harris corner detector \cite{harris1988combined}.

@@

For a thorough review of the history of feature detectors, Tuytelaars 
and Mikolajczyk presents an excellent retrospect in 
\cite{tuytelaars2008local}.

@@

Instead of using the pixel values, we can create a histogram of the 
gradients surrounding the feature point as Lowe did in his Scale 
Invariant Feature Detector (SIFT) \cite{lowe2004sift}. The histogram he 
proposes is based on the gradient locations and orientations weighted by
the gradient magnitude. To make sure that the histogram is rotation 
invariant, the bins of the histogram are normalized and start with the 
bin of the highest magnitude. The SIFT descriptor has been very widely 
used since its introduction and has become a de facto standard in 
computer vision. To enhance the computational time needed to compute 
descriptors, Bay et al.\ introduced the Speeded Up Robust Features 
(SURF) that rely on a histogram of wavelets instead of gradients as used 
in SIFT \cite{bay2006surf}.

A radically different approach to a feature descriptor is to measure the
intensity difference between a number of pixels surrounding the key 
point and simply assign the bits to a vector that is returned as the 
feature descriptor. Calonder et al.\ \cite{calonder2010brief} introduced 
a feature descriptor, Binary Robust Independent Elementary Features 
(BRIEF), based on this idea. The main advantage of their approach is 
reduced processing and memory needed to create, store and match the 
feature descriptors. Interestingly the positions used for pixel wise 
comparisons were selected at random and tested against similar random 
configurations. They then went on to use the configuration yielding the 
best result.  The simplicity of this method allows for real time 
matching of objects from a video stream, something that would be too 
computationally tolling when using a descriptor like SIFT\@. BRIEF 
however fails to be rotation invariant making it less useful in cases 
where images are not necessarily upright.

@@

Most applications of feature points use them to match features across 
different images. The standard way most feature points are matched 
involve comparing the feature descriptor using some kind of distance 
measure. If we match all features of one image with all features of 
another and discard all matches that are below a certain threshold of 
similarity, we have in essence a set of matches between the
two images. For most feature points (for example SIFT and SURF), this 
similarity is measured by the cosine similarity. Binary descriptors such
as BRIEF can take advantage of the faster computation of the hamming 
distance.

@@

An alternative is to pick out different areas in each image and pair 
areas when correspondence pairs are mainly found going from one area to 
another. This allows for the filtering of all correspondences that do 
not fall within paired areas. Examples include \emph{Isomatch} 
\cite{das2008event}, which clusters feature points according to their 
position in the image, and the method from \cite{wu2011robust}, which 
uses maximally stable extremal regions (MSER) feature points to 
designate areas.

@@

In practice there are many situations where relying on the geometry of 
the image to filter correspondences is not possible, for example images 
without much overlap, or adjacent objects in one image are that are 
separated in another. In these cases the assumptions behind the 
geometric algorithms will often cause them to return incorrect matches.

@@

In the absence of geometric constraints, only few methods are available.  
As pointed out in \cite{szeliski2010} and mentioned above, the simplest 
approach to matching feature points is to return all matches above a 
given threshold of similarity.  This can be done with a fixed threshold 
or by finding the nearest neighbors for each point (still filtering by a 
threshold to avoid including matches for feature points that have no 
counterpart in the other image).  When Lowe \cite{lowe2004sift} proposed 
the scale-invariant feature transform he included an alternative 
matching measure. Instead of rating matching by similarity assessed the 
uniqueness of a given match by looking at the two nearest neighbors of 
each feature point and calculating the matching score as the ratio of 
similarities. This method will later be referred to as \emph{Ratio}.  
Ranking the scores by their uniqueness and picking the $n$ best produces 
a set of correspondences that are distinctly matched across two images.  
Of these three methods (fixed threshold, nearest neighbors, and 
\emph{Ratio}), \emph{Ratio} was found to perform best by a slight margin 
\cite{mikolajczyk2005performance}.
